---
title: "Data Mining Final Project"
output:
  pdf_document: default
  word_document: default
---


**Your Name**: Anasa Alamgir
**Your G Number**: G01300460


```{r warning = FALSE, message = FALSE, include=FALSE}
# Suppress dplyr summarize grouping warning messages
options(dplyr.summarise.inform = FALSE)
## Add R libraries here
library(tidyverse)
library(tidymodels)
library(discrim)
library(kknn)
library(dplyr)
library(klaR)
library(ggplot2)
library(vip)
library(magrittr)
# Load data
loans_df <- read_rds("loan_data.rds")
```

# Introduction 

The National Bank has experienced customers defaulting on their loans 
and therefore seeks to predict whether an applicant will default on their loan
in order to protect the Bank from large financial losses.This project aims to 
explore the factors that lead to loan default and use machine learning models 
to predict the chance of an applicant defaulting on their loan in the future. 


The loans data set contains information on 3 and 5 year loans that were originated
in 2017 from customers residing in the Middle Atlantic and Northeast regions
of the United States. Exploratory data analysis can help to find the relationship 
between whether the applicant defaults and the various factors that affect them 
defaulting on the loan.


# Data

The loans data set shows applicant data with information on the loan amount, 
installment amount, interest rate, loan purpose, application type, loan 
term, home ownership, annual income, current job years, debt to income
ratio, total credit line, years of credit history, history of 
bankruptcy and history of tax liens.

In the exploratory data analysis, this report will be using these 
factors:
loan amount, installments, interest rates, loan purpose, application
type, term, home ownership, annual income, current job years, 
debt to income ratio, and years of credit history.

Some questions this report intends to answer:

- What are the factors related to applicants defaulting on their loan?
- Is it possible to predict whether a customer will default on their loan?
  - How accurate are these predictions?
  - How many errors is the model expected to produce?
- Are there any policies that the bank can implement to reduce the risk of
loan default?


# Summary of Results

The default response variable in this data frame is loan_default, which records
whether an applicant has defaulted or not. This variable has also been coded with 
'Yes' and 'No' factors. Therefore using visualization techniques this report will 
show which other factors can explain why some applicants default and others do not.


The data visualization results show that a lower loan amount defaulted more often,
and applicants with lower median income defaulted on their loans as well. 
In terms of home ownership, renters defaulted more often than applicants that 
owned their home or paid mortgage on their home. Applicants with a higher 
debt ratio also defaulted since they have more debt that what they earn. Besides
that, a higher interest rate also led to applicants defaulting on their loans. 
The predictive modelling section elaborates on the factors that significantly
affect loan default and the steps that the bank might find beneficial to 
implement to reduce the risk of loan default.



# Predictive Modeling 

In order to find the most significant factors affecting applicants to default on 
their loans, this report implements Logistic Regression, Linear Discriminant 
Analysis and K-Nearest Neighbors algorithms. It also uses ROC and AUC to measure
the accuracy of the model. The most accurate model will be used to find the
most significant factors affecting loan default. 

The ROC AUC of the Logistic Regression model is 99%, which is the highest of 
all the three models created. Based on the results of the vip function,
there are four significant factors affecting loan default:

- interest rate
- term: five year
- installment
- loan amount

Among these four, interest rate and five year term are the most important
factors. The model also predicts 8.1% of errors that might be made by the model
(negative predictions that are actually positive). Applicant seem to be most 
affected by interest rates and loan term, therefore in order to reduce the risk
of loan default, the bank can take these steps:
 
- Reduce the loan term so that applicants can pay off their loan earlier
- Reduce the loan interest rate so that applicants can pay off the entire loan
before it falls into default. 

```{r, echo=FALSE, include=FALSE}
#split loans_df into training and test sets
set.seed(172)
loan_split <- initial_split(loans_df, prop = 0.75,
                            strata = loan_default)
loan_training <- loan_split %>% training()
loan_test <- loan_split %>% testing()

#cross validation folds for hyperparameter tuning
set.seed(172) 
loan_folds <- vfold_cv(loan_training, v = 6)

#feature engineering
loan_recipe <- recipe(loan_default ~ .,data = loan_training) %>%
  step_YeoJohnson(all_numeric(), -all_outcomes()) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes())
loan_recipe %>%
  prep(training = loan_training) %>%
  bake(new_data = NULL)

#specify model
loan_logistic <- logistic_reg() %>%
  set_engine('glm') %>%
  set_mode('classification')

#create workflow
logistic_wf <- workflow() %>%
  add_model(loan_logistic) %>%
  add_recipe(loan_recipe)
#roc curve and auc
logistic_fit <- logistic_wf %>% last_fit(split = loan_split)
# collect predictions
logistic_predictions <- logistic_fit %>% collect_predictions()
```

### Confusion Matrix for Logistic regression model
```{r, echo=FALSE}
#confusion matrix
conf_mat(logistic_predictions,
         truth = loan_default,
         estimate = .pred_class)
```


### ROC AUC for regression model
```{r, echo=FALSE}
#area under curve
roc_auc(logistic_predictions, truth = loan_default, .pred_yes)
```






# Exploratory Data Analysis 


## Question 1

**Question**: Is there a relationship between loan default and loan amount?

**Answer**: The distribution of loans that defaulted based on the loan amount 
shows that a greater number of loans that defaulted had a smaller loan amount.
There are a lower number of loans which defaulted at a higher loan amount.


## Question 2

**Question**: Is there a relationship between loan default and history of missed
payments in the past 2 years?

**Answer**: The summary does not show a positive relationship between history
of missed payments and whether the applicant defaulted. There is a higher
number of loans that defaulted but the applicants did not have history of missed
payments in the past 2 years. On the other hand, there is a much higher number 
of loans that did not default but the applicants had history of missed payments 
in the past 2 years.


## Question 3

**Question**: Is there a relationship between loan default and annual income?

**Answer**: Yes, the summary clearly shows that applicants with a lower
median income ($60,000) and lower average income ($67,818.80) had defaulted
on their loans compared to applicants with higher average and median incomes.


## Question 4

**Question**: How does loan default relate to home ownership?

**Answer**: The data indicates that applicants renting their home are more
likely to default on their loans compared to a far lesser amount of 
applicants that own their home.



## Question 5

**Question**: Is there a relationship between loan default and debt-to-income 
ratio?

**Answer**: The density plot shows that applicants with higher debt-to-income
ratio defaulted more often than applicants with a lower debt to income ratio. 
This is most likely because applicants with higher debt to income ratio already
have substantial amount of debt compared to their income to be able to pay
on their installments. Applicants with much lower debt to income ratio are
less likely to have defaulted on their loans.



## Question 6

**Question**: How is loan default related to application type and 
years of credit history?

**Answer**: The summary indicates that individual applications that got 
defaulted had a smaller average credit history when compared to joint applications
with lower credit history that went into default. On the other hand, for both
types of applications, the ones with higher average credit history did not go 
into default.



## Question 7

**Question**: Is there a relationship between loan default rate and interest 
rates and loan purpose?

**Answer**: Out of the applicants that defaulted on their loans, small business
and credit card loans have the highest median interest rates. Therefore, it can 
be stated that applicants are more likely to default on their loans with a higher
interest rate, especially when it is for a small business or credit card.






# Appendix


## Tables and visualizations

### question 1: Is there a relationship between loan default and loan amount?
```{r, fig.asp=0.6, fig.width=8, echo=FALSE}
# question 1: Is there a relationship between loan default and loan amount?
# visualization of loans defaulted based on loan amount
loans_df %>% filter(loan_default=="yes") %>% ggplot(
  aes(x = loan_amount)
) +
  geom_histogram(bins=12, color="white", fill="#E64D54") +
  theme_light() +
  labs(
    title = "Distribution of loans that defaulted based on loan amount",
    x = "loan amount",
    y = "number of loans"
  )
```


### question 2: Is there a relationship between loan default and history of missed payments in the past 2 years?
```{r, echo=FALSE}
# question 2: Is there a relationship between loan default and history of missed
# payments in the past 2 years?
# summary of missed payments in the past two years grouped by loan default
loans_df %>%
  group_by(loan_default, missed_payment_2_yr) %>%
  summarise(
    num_loans = n()
  )
```

### question 3: Is there a relationship between loan default and annual income?
```{r, echo=FALSE}
#question 3: Is there a relationship between loan default and annual income?
# summary of annual income by loan default
loans_df %>% group_by(loan_default) %>%
  summarise(
    avg_income = mean(annual_income),
    median_income = median(annual_income),
    min_income = min(annual_income),
    sd_income = sd(annual_income)
  )
```


### question 4: How does loan default relate to home ownership?
```{r, echo=FALSE}
#question 4: How does loan default relate to home ownership?
#visualization of loans based on home ownership
loans_df %>% filter(loan_default=="yes") %>% 
  ggplot(aes(x=homeownership, fill=loan_default)) + 
  geom_bar(fill="#7ACBFF") +
  theme_minimal() + labs(
    title="Distribution of loans defaulted based on home ownership", 
    fill="loan default"
  )
```


### question 5:  Is there a relationship between loan default and debt-to-income ratio?
```{r, fig.width=9, fig.asp=0.5, echo=FALSE}
#question 5:  Is there a relationship between loan default and debt-to-income 
#ratio?
# visualization of the distribution of debt to income ratio by loans defaulted
loans_df %>% ggplot(
  aes(x = debt_to_income, color = loan_default)
) + geom_density() + xlim(0,100)+ 
  theme_light() +
scale_fill_brewer(palette="Set1") +
  labs(
    title = "Distribution of debt to income ratio by loans defaulted",
    color = "loan defaulted",
    x = "debt to income ratio"
  )
```


### question 6: How is loan default related to application type and years of credit history?
```{r, echo=FALSE}
# question 6: How is loan default related to application type and 
# years of credit history?
# summary of years of credit history based on application type and loan default
loans_df %>%  group_by(loan_default,application_type) %>% summarize(
  avg_credit_history=mean(years_credit_history),
  median_cred_history=median(years_credit_history),
  sd_cred_history=sd(years_credit_history)
  )
```


### question 7: Is there a relationship between loan default rate and interest rates and loan purpose?
```{r, echo=FALSE}
# question 7: Is there a relationship between loan default rate and interest 
# rates and loan purpose?
# summary of interest rates based on loan purpose and loan default
loans_df %>% filter(loan_default == "yes") %>%
  group_by(loan_default, loan_purpose) %>%
  summarise(
    med_interestrate = median(interest_rate))
```


## Predictive models 


### Model 1 Logistic Regression

```{r}
#split loans_df into training and test sets
set.seed(172)
loan_split <- initial_split(loans_df, prop = 0.75,
                            strata = loan_default)
loan_training <- loan_split %>% training()
loan_test <- loan_split %>% testing()
```

```{r}
#cross validation folds for hyperparameter tuning
set.seed(172) 
loan_folds <- vfold_cv(loan_training, v = 6)
```

```{r}
#feature engineering
loan_recipe <- recipe(loan_default ~ .,data = loan_training) %>%
  step_YeoJohnson(all_numeric(), -all_outcomes()) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes())
loan_recipe %>%
  prep(training = loan_training) %>%
  bake(new_data = NULL)
```

```{r}
#specify model
loan_logistic <- logistic_reg() %>%
  set_engine('glm') %>%
  set_mode('classification')
```

```{r}
#create workflow
logistic_wf <- workflow() %>%
  add_model(loan_logistic) %>%
  add_recipe(loan_recipe)
#roc curve and auc
logistic_fit <- logistic_wf %>% last_fit(split = loan_split)
# collect predictions
logistic_predictions <- logistic_fit %>% collect_predictions()
```

```{r}
#confusion matrix
conf_mat(logistic_predictions,
         truth = loan_default,
         estimate = .pred_class)
```

```{r}
#roc curve
roc_curve(logistic_predictions, truth = loan_default, estimate = .pred_yes) %>% 
  autoplot()
```

```{r}
#area under curve
roc_auc(logistic_predictions, truth = loan_default, .pred_yes)
```
Accuracy: 99%


```{r}
#model summary 
log_model <- glm(loan_default ~., data = loan_training, family = binomial())
tidy(log_model)
```

```{r}
#get summary
summary(log_model)
```


```{r}
# get vip
vip(log_model)
```



### Model 2 LDA

```{r}
#specify model
loan_lda <- discrim_regularized(frac_common_cov = 1) %>%
  set_engine("klaR") %>%
  set_mode("classification")
```


```{r}
#workflow
lda_wf <- workflow() %>% add_model(loan_lda) %>% add_recipe(loan_recipe)
```

```{r}
#fit workflow
lda_fit <- lda_wf %>% last_fit(split=loan_split)
```

```{r}
#collect predictions
lda_predictions <- lda_fit %>% collect_predictions()
```

```{r}
#confusion matrix
conf_mat(lda_predictions, truth = loan_default, estimate = .pred_class)
```

```{r}
#roc curve
roc_curve(lda_predictions, truth=loan_default, estimate= .pred_yes) %>%
  autoplot()
```

```{r}
#Area under ROC
roc_auc(lda_predictions, truth=loan_default, .pred_yes)
```
Accuracy: 98.99%



### Model 3 K-Nearest Neighbor 

```{r}
#specify model
knn_model <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")
```

```{r}
#workflow
knn_wf <- workflow() %>%
  add_model(knn_model) %>%
  add_recipe(loan_recipe)
```

```{r}
#create grid for hyperparameter testing
k_grid <- tibble(neighbors = c(10,20,30,40,50,75,100,125,150))
```

```{r}
#tuning wf
set.seed(271)
knn_tuning <- knn_wf %>% tune_grid(resamples=loan_folds, grid=k_grid)
#select best model from tuning result
best_k <- knn_tuning %>% select_best(metric='roc_auc')
#add optimal model to wf
final_knn_wf <- knn_wf %>% finalize_workflow(best_k)
#fit model
knn_fit <- final_knn_wf %>% last_fit(split=loan_split)
```

```{r}
#get df of test prediction results
knn_predictions <- knn_fit %>% collect_predictions()
```

```{r}
#confusion matrix
conf_mat(knn_predictions, truth= loan_default, estimate = .pred_class)
```

```{r}
#roc curve 
roc_curve(knn_predictions, truth = loan_default, estimate= .pred_yes) %>%
  autoplot()

```

```{r}
#Are under ROC
roc_auc(knn_predictions, truth = loan_default, estimate= .pred_yes)
```

Accuracy: 90%




--- End of the Project ---